---
title: "About This Website"
---

This is an [R Markdown website](http://rmarkdown.rstudio.com/rmarkdown_websites.html), created by [Julia Wrobel](http://juliawrobel.com) for a single-cell spatial proteomics and multiplex imaging data workshop at the 2024 [Pacific Symposium on Biocomputing (PSB)](https://psb.stanford.edu/). For raw materials, go to the [GitHub repository for this website](https://github.com/julia-wrobel/PSB_scProteomics).

In the short course, we will use several packages which are listed below. Code to install the packages follows, and version information about each of these packages and their dependencies is at the bottom of this page.

Installing packages from CRAN

```{r, eval=FALSE}
# main packages
install.packages(c("tidyverse", "spatstat", "spatialTIME", "mxnorm","survival", "scSpatialSim"))

# additional packages used for visualization
install.packages(c("viridis", "patchwork"))
```

Installing packages from GitHub

```{r, eval = FALSE}
# install devtools
install.packages("devtools")

# mxfda
devtools::install_github("julia-wrobel/mxfda")
```



Installing packages from Bioconductor

```{r, eval = FALSE}
# install Bioconductor
BiocManager::install(version = "3.16")

# VectraPolarisData
BiocManager::install("VectraPolarisData")
```



Versions of key packages:

```{r}
devtools::session_info(c("tidyverse", "VectraPolarisData", "spatstat", "spatialTIME", "mxfda"))
```

