---
title: "Introduction to multiplex single-cell imaging data"
output:
  html_document: 
    toc: true
    toc_float: true
---



```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%",
  message = FALSE
)

```




## Slide Deck

<iframe class="speakerdeck-iframe" frameborder="0" src="https://speakerdeck.com/player/19588e79e83146a8bc844ed2f335c79e" title="MI_intro" allowfullscreen="true" style="border: 0px; background: padding-box padding-box rgba(0, 0, 0, 0.1); margin: 0px; padding: 0px; border-radius: 6px; box-shadow: rgba(0, 0, 0, 0.2) 0px 5px 40px; width: 100%; height: auto; aspect-ratio: 560 / 314;" data-ratio="1.78343949044586"></iframe>


## Overview

In this module I briefly discuss loading and visualizing raw TIFF data. The rest of the workshop will focus on data in tabular form post segmentation.

### Lung cancer data

The following image is a single ROI from a tissue slice from a patient with non small cell lung carcinoma.Cell and tissue segmentation and cell phenotyping were performed using [inForm](https://www.akoyabio.com/phenoimager/software/inform-tissue-finder/) software. For each image, there are multiple Tiff files output by inForm:


* `lung_component_data.tif`: the multichannel file we are most interested in
* `lung_binary_seg_maps.tif`: 3 channel file with binary segmentation maps
* `lung.tif`: A composite image in RGB format. 

You can download these files from the Data tab and store them in a local directory. 

```{r}

#BiocManager::install("EBImage")

# load relevant libraries
library(tidyverse)
library(tiff)
library(EBImage)


# define your specific file path
my_path = "./Data/"

# define file for different .tif files
component = paste0(my_path, "lung_component_data.tif")
binseg = paste0(my_path, "lung_binary_seg_maps.tif")
composite_tif = paste0(my_path, "lung.tif")
```


#### Component file


The `component.tif` file contains the multichannel tiff, which is a stack 0f individual images (one for each marker). Below I define a function to read the component file and extract metadata.

```{r}
# this function is from phenoptr package on GitHub
# extracts metadata specific to Vectra3/VectraPolaris images
read_components <- function(path) {
  stopifnot(file.exists(path), endsWith(path, 'component_data.tif'))

  # The readTIFF documentation for `as.is`` is misleading.
  # To read actual float values for components, we want `as.is=FALSE`.
  tif = tiff::readTIFF(path, all=TRUE, info=TRUE)

  # Get the image descriptions and figure out which ones are components
  infos = purrr::map_chr(tif, ~attr(., 'description'))
  images = grepl('FullResolution', infos)

  # Get the component names
  names = stringr::str_match(infos[images], '<Name>(.*)</Name>')[, 2]

  purrr::set_names(tif[images], names)
}

image_stack = read_components(component)
```

The `image_stack` object is now a list, where each list element is a different marker

```{r}
map(image_stack, dim)
```


I can use the `display()` function from the `EBImage` package to plot an individual channel. Below the DAPI (nucleus) channel is plotted.

```{r}
dapi_channel = t(image_stack$"DAPI (DAPI)")
EBImage::display(dapi_channel, method = "raster")


# plot the image in a specified color
dapi = rgbImage(blue=.1*dapi_channel, red = .02 * dapi_channel)
display(dapi, method = "raster")
```

#### Segmentation files

This data was collected on a Vectra 3 instrument and segmented using [inForm](https://www.akoyabio.com/phenoimager/software/inform-tissue-finder/) tissue analysis     software from Akoya Biosciences. This is proprietary image analysis software that can perform cell and tissue segmentation and cell phenotyping. The segmentation below were performed by that software.

```{r}
# source phenoptr code for loading file and metadata, based on tiff::readTIFF.
source("https://raw.githubusercontent.com/akoyabio/phenoptr/main/R/read_maps.R")

segmentations = read_maps(binseg)
names(segmentations)
```

Nucleus segmentation

```{r}
nucleus = segmentations[['Membrane']]
plot(as.raster(nucleus, max = max(nucleus)))
```


Tissue segmentation

```{r}
tissue = segmentations[['Tissue']]

plot(as.raster(tissue, max = max(tissue)))
```

#### Composite file

```{r}
# source phenoptr code for loading file and metadata, based on tiff::readTIFF.
source("https://raw.githubusercontent.com/akoyabio/phenoptr/main/R/read_composites.R")

composite = read_composites(composite_tif)
```

```{r}
composite = as.raster(composite[[1]])
plot(composite)
```



## References

Below are original citations for the two example datasets I will be working with in this tutorial.

* [Paper for ovarian cancer data](https://aacrjournals.org/mcr/article/19/12/1973/675069/The-Spatial-Context-of-Tumor-Infiltrating-Immune)
* [Paper for non-small cell lung carcinoma data](https://pubmed.ncbi.nlm.nih.gov/34048945/)

Resources from Akoya Biosciences on working with this data

* [phenoptr R package](https://akoyabio.github.io/phenoptr/)

Below are some references that review this rapidly growing area.

* [textbook chapter, Wrobel and Vandekar](http://juliawrobel.com/Downloads/mIF_chapter.pdf)
* [Challenges and Opportunities in the Statistical Analysis of Multiplex Immunofluorescence Data](https://www.mdpi.com/2072-6694/13/12/3031)
* [Catching up with multiplexed tissue imaging](https://www.nature.com/articles/s41592-022-01428-z) 
* [The emerging landscape of spatial profiling technologies](https://www.nature.com/articles/s41576-022-00515-3)