---
title: "Spatial analysis of multiplex imaging data"
output:
  html_document: 
    toc: true
    toc_float: true
---

```{r, echo = FALSE, message = FALSE}
library(tidyverse)
library(patchwork)


knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .6,
  message = FALSE,
  out.width = "90%",
  warning = FALSE
)

theme_set(theme_bw() + theme(legend.position = "bottom"))
```




## Slide Deck

<iframe class="speakerdeck-iframe" frameborder="0" src="https://speakerdeck.com/player/034999f5e7844fb9a3e778d055b3bedc" title="MI_spatialAnalysis" allowfullscreen="true" style="border: 0px; background: padding-box padding-box rgba(0, 0, 0, 0.1); margin: 0px; padding: 0px; border-radius: 6px; box-shadow: rgba(0, 0, 0, 0.2) 0px 5px 40px; width: 100%; height: auto; aspect-ratio: 560 / 314;" data-ratio="1.78343949044586"></iframe>

## Overview

This module focuses on extracting spatial summary metrics from multiplex imaging data.

Load data and libraries.

```{r}
library(tidyverse)
library(patchwork)

# load processed ovarian cancer data
load(url("https://github.com/julia-wrobel/MI_tutorial/raw/main/Data/ovarian.RDA"))
```


### Visualize point patterns

Below we visualize a point pattern for a single subject from the ovarian cancer dataset.

* pink: stroma area
* light green: tumor area
* dark green: macrophages
* purple: B-cells

```{r point_pattern}

# subset to a single subject
subj_df = filter(ovarian_df, sample_id == 7) 

# 7, 110, 34
# plot macrophage and B-cells
all = subj_df %>%
  ggplot(aes(x, y)) +
  geom_point(aes(color = tissue_category),
             size = 0.1, alpha = 0.7) +
  scale_color_manual(values = c("pink", "lightgreen")) +
  geom_point(data = filter(subj_df,  phenotype_cd68 == "CD68+"), 
             color = "darkgreen", size = .5, shape = 4) +
    geom_point(data = filter(subj_df,  phenotype_cd19 == "CD19+"), 
             color = "purple", size = .5, shape = 6) +
  theme(legend.position ="bottom")
  
cell_subset = subj_df %>%
  ggplot(aes(x, y)) +
  geom_point(data = filter(subj_df,  phenotype_cd68 == "CD68+"), 
             color = "darkgreen", size = .5, shape = 4) +
    geom_point(data = filter(subj_df,  phenotype_cd19 == "CD19+"), 
             color = "purple", size = .5, shape = 6) +
  theme(legend.position ="bottom")

all / (cell_subset + all)
```

## Analysis using spatstat package

The `spatstat` package in `R` has great resources for analyzing spatial point patterns. Let's use the image above to do some exploratory analysis with spatstat.

We will do spatial analysis on the distribution of B-cells and macrophages, so we subset to only these cell types, and create a `phenotype` variable that designates whether a cell is a B cell or a macrophage. Then we create a `ppp` object, which is what the `spatstat` package uses for storage and analysis of point pattern data.

```{r}
library(spatstat)

subj_df = subj_df %>%
  # create a phenotype variable that is B-cell, macrophage, or other
  mutate(phenotype = case_when(
    phenotype_cd68 == "CD68+" ~ "macrophage",
    phenotype_cd19 == "CD19+" ~ "B-cell",
    TRUE ~ "other"
  )) %>%
  filter(phenotype != "other") %>%
  select(x,y, immune, tissue_category, phenotype)

# first define window of observation for your image
w = convexhull.xy(subj_df$x,subj_df$y)

# create ppp object as multitype point pattern
# need to factor the marks variable for ppp object to be treated as a multitype point pattern
ovarian_pp = ppp(subj_df$x,subj_df$y, window = w,
                 marks = factor(subj_df$phenotype))
```


### Univariate Ripley's  K


First we calculate Ripley's K for B cells.  


```{r}
k_bcell =  Kest(subset(ovarian_pp, marks == "B-cell"), correction = "isotropic")
```

We can plot this `k_bcell` object using `spatstat` to visualize the estimated K ($\hat{K}^{iso}$) compared to the K under CSR ($K^{pois}$).

```{r}
plot(k_bcell)
```

Now visualize the K-function for macrophages in this image:

```{r}
k_mac =  Kest(subset(ovarian_pp, marks == "macrophage"), 
              correction = "isotropic")
plot(k_mac)
```

It appears that there is some degree of clustering for both B-cells and macrophages in this image. Are these statistically significant?

We can test this using the `envelope` function, which performs Monte Carlo simulations of the K function under CSR. Plot below shows pointwise envelope for B-cell K-function. Pointwise means that these are performed separately for each value of *r*. Can only be interpreted if a specific value of *r* is chosen in advance.

```{r}

e_test = envelope(subset(ovarian_pp, marks == "B-cell"), 
               Kest, correction = "isotropic", global = TRUE)

plot(e_test)

```

Points outside the envelope indicate significant departure from CSR. Note that this could be due to inhomogeneity in the distribution of cells, or due to clustering. See `spatstat::Kinhom` function for a test that takes into account inhomogeneity. See `spatstat::Lest` function to compute the L-function.


### Bivariate Ripley's  K

The bivariate K function measures whether the spatial distribution of 2 cell types is independent. Deviation from the $K^{pois}$ line suggests a spatial relationship between the B-cells and macrophages in the image.

```{r}
k_biv = Kcross(ovarian_pp, "B-cell", "macrophage", correction = "isotropic")

plot(k_biv)
```

### Nearest neighbor G function

Can interpret as probability of a neighboring cell occurring within radius r. 


```{r}
G_bcell = Gest(subset(ovarian_pp, marks == "B-cell"), correction = "rs")

plot(G_bcell)
```

Its bivariate counterpart 

```{r}
G_biv = Gcross(ovarian_pp, correction = "rs")

plot(G_biv)


e_test = envelope(ovarian_pp, 
               Gcross,  global = TRUE)

plot(e_test)

```

## SpatialTIME package


Organize data for use in the `spatialTIME` package.

```{r}
library(spatialTIME)

# subsetting to 2 subjects for faster computation and ease of visualization
ids = c( 7, 128)

spatial_ls = filter(ovarian_df, sample_id %in% ids) %>%
    mutate(XMax = x, XMin = x,
         YMax = y, YMin = y,
         sample_id = factor(sample_id),
         id = sample_id,
         cd19 = ifelse(phenotype_cd19 == "CD19+", 1, 0),
         cd3 = ifelse(phenotype_cd3 == "CD3+", 1, 0),
         cd8 = ifelse(phenotype_cd8 == "CD8+", 1, 0),
         cd68 = ifelse(phenotype_cd68 == "CD68+", 1, 0)) %>%
  dplyr::select(sample_id, id, XMax, XMin, YMax, YMin, 
                cd19, cd3, cd8, cd68, tissue_category) %>%
  nest_by(id)

# define clinical data
clinical_data = ovarian_df %>%
  #filter(ovarian_df, sample_id %in% ids) %>%
  dplyr::select(sample_id, survival_time, death, stage_bin, BRCA_mutation) %>%
  distinct() %>%
  mutate(id = sample_id)

sample_data = ovarian_df %>%
  group_by(sample_id) %>%
  summarize(total_cells = n()) %>%
  ungroup() %>%
  mutate(id = sample_id)

  
df_spatialTIME = spatialTIME::create_mif(
  clinical_data = clinical_data,
  sample_data = sample_data,
  spatial_list = spatial_ls$data,
  patient_id = "id",
  sample_id = "sample_id"
           )

rm(spatial_ls)
```

Plotting samples

```{r}

cell_types = c("cd68", "cd19")
values =  RColorBrewer::brewer.pal(length(cell_types), "Accent")


#add an element in the `derived` object position
df_spatialTIME <- plot_immunoflo(df_spatialTIME, 
                   plot_title = "sample_id",  mnames = cell_types,
                   cell_type = "tissue_category")



```


Calculate spatial summary metrics (calculating univariate K function across samples and phenotypes).

```{r}
df_spatialTIME <- ripleys_k(mif = df_spatialTIME, 
                            mnames = cell_types, 
                            method = "K",
                            r_range = seq(0, 300, 10),
                            edge_correction = 'translation'
                            permute = TRUE,
                            num_permutations = 100
                            )

```

Plot images(top row) and  K functions (bottom row).

```{r, fig.width= 12, fig.length = 8}

k_funs = df_spatialTIME$derived$univariate_Count  %>%
  janitor::clean_names() %>%
  ggplot(aes(x = r, y = observed_k)) +
  geom_line(aes(group = marker, 
             color = marker)) +
  geom_line(aes(r, theoretical_csr), linetype = 2) +
  facet_wrap(~sample_id, scales = 'free', nrow = 1) + theme_bw() +
  theme(legend.position = "bottom") +
  scale_color_manual(values = values)


imgs = patchwork::wrap_plots(df_spatialTIME$derived$spatial_plots, nrow = 1)

imgs / k_funs

```


Also allows for calculation of bivariate K, univariate and bivariate L, G, with and without permutations.

## References

* [Textbook for spatstat package](https://spatstat.org/book.html)
* [spatstat JSS article](https://www.jstatsoft.org/article/view/v012i06)
* [Paper for ovarian cancer data](https://aacrjournals.org/mcr/article/19/12/1973/675069/The-Spatial-Context-of-Tumor-Infiltrating-Immune)
* [paper for spatialTIME package](https://academic.oup.com/bioinformatics/article/37/23/4584/6420699) and associated [methods paper](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1009900)
* [textbook chapter, Wrobel and Vandekar](https://link.springer.com/protocol/10.1007/978-1-0716-2986-4_8)
* [Challenges and Opportunities in the Statistical Analysis of Multiplex Immunofluorescence Data](https://www.mdpi.com/2072-6694/13/12/3031)
* [Chervoneva, 2022](https://academic.oup.com/bioinformatics/article/37/10/1452/6020255)

