---
title: "VectraPolarisData: experiment-scale multiplex imaging data"
output:
  html_document: 
    code_folding: hide
    toc: true
    toc_float: true
--- 



```{r, echo = FALSE, message = FALSE}
library(tidyverse)

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_bw() + theme(legend.position = "bottom"))
```




## Slide Deck

<iframe class="speakerdeck-iframe" frameborder="0" src="https://speakerdeck.com/player/5573ed99b24a464885bbbab47af138d5" title="MI_VectraPolarisData" allowfullscreen="true" style="border: 0px; background: padding-box padding-box rgba(0, 0, 0, 0.1); margin: 0px; padding: 0px; border-radius: 6px; box-shadow: rgba(0, 0, 0, 0.2) 0px 5px 40px; width: 100%; height: auto; aspect-ratio: 560 / 314;" data-ratio="1.78343949044586"></iframe>

## Overview


The [VectraPolarisData](https://bioconductor.org/packages/release/data/experiment/html/VectraPolarisData.html) [ExperimentHub](http://bioconductor.org/packages/release/bioc/html/ExperimentHub.html) package provides two large multiplex immunofluorescence datasets collected using Akoya Biosciences Vectra 3 and Vectra Polaris platforms. Image preprocessing (cell segmentation and phenotyping) was performed using [inForm](https://www.akoyabio.com/phenoimager/software/inform-tissue-finder/) software. Data are provided as objects of class [SpatialExperiment](https://bioconductor.org/packages/release/bioc/html/SpatialExperiment.html).

```{r, message = FALSE}
# load libraries
library(tidyverse)
```


## Install and load data

The data is publicly available on Bioconductor as the package `VectraPolarisData`. You can install the package from Bioconductor here:

```{r, eval = FALSE, echo = TRUE, message = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
     install.packages("BiocManager")
}

BiocManager::install("VectraPolarisData")

```


## Ovarian cancer data 

Load the high grade serous ovarian cancer data. This dataset has been segmented and phenotyped, and has one image per patient.

```{r, message = FALSE}
library(VectraPolarisData)

# load spatial experiment object
oc <- HumanOvarianCancerVP()
```

Inspect the `SpatialExperiment` object.

```{r}
# check object
oc
```

The object stores the following marker values for each marker and each cell

* `intensities`: mean marker intensity for entire cell
* `nucleus_intensities`: mean marker intensity in the nucleus
* `membrane_intensities`: mean marker intensity in the cell membrane

```{r}
assayNames(oc)
```


Row names provide the names of the markers.

```{r}
rownames(oc)
```


The colData slot contains cell-level covariates like cell phenotype, cell area, and, min, max, and standard deviation of intensity values for each marker.  

```{r}
dim(colData(oc))

```


Many of the variables are summary statistics of marker intensity values. More information on the different variables in this and the lung cancer dataset can be found in the [package vignette](https://bioconductor.org/packages/release/data/experiment/vignettes/VectraPolarisData/inst/doc/VectraPolarisData.html).

```{r}
set.seed(12333)
sample(names(colData(oc)), 15)
```


### Converting to dataframe 

Storing the data as a `SpatialExperiment` allows for analysis using tools that are built into the `SpatialExperiment` workflow.  However, it is often useful to work with the data as a `data.frame` object instead.

Code below converts the ovarian cancer dataset from a `SpatialExperiment` object to a `data.frame`. We perform some data cleaning steps as well.

```{r}

## Assays slots
assays_slot <- assays(oc)
intensities_df <- assays_slot$intensities
nucleus_intensities_df<- assays_slot$nucleus_intensities
rownames(nucleus_intensities_df) <- paste0("nucleus_", rownames(nucleus_intensities_df))
membrane_intensities_df<- assays_slot$membrane_intensities
rownames(membrane_intensities_df) <- paste0("membrane_", rownames(membrane_intensities_df))

# colData and spatialData
colData_df <- colData(oc)
spatialCoords_df <- spatialCoords(oc)

# clinical data
patient_level_ovarian <- metadata(oc)$clinical_data %>%
  # create binary stage variable
  dplyr::mutate(stage_bin = ifelse(stage %in% c("1", "2"), 0, 1))

# only include samples for whom we have an id in the patient dataset, which is 128- the other 4 are controls
# give Markers shorter names
cell_level_ovarian <- as.data.frame(cbind(colData_df, 
                                     spatialCoords_df,
                                     t(intensities_df),
                                     t(nucleus_intensities_df),
                                     t(membrane_intensities_df))
                               ) %>%
  dplyr::rename(cd19 = cd19_opal_480,
                cd68 = cd68_opal_520,
                cd3 = cd3_opal_540, 
                cd8 = cd8_opal_650,
                ier3 = ier3_opal_620,
                pstat3 = p_stat3_opal_570,
                ck = ck_opal_780,
                ki67 = ki67_opal_690) %>%
  # define cell type 'immune'
  mutate(immune = ifelse(phenotype_cd19 == "CD19+" | phenotype_cd8 == "CD8+" |
              phenotype_cd3 == "CD3+" | phenotype_cd68 == "CD68+", "immune", "other")) %>%
  dplyr::select(contains("id"), tissue_category, contains("phenotype"),
                contains("position"), ck:dapi, immune) %>%
  # only retain 128 subjects who have clinical data (other 4 are controls)
  dplyr::filter(sample_id %in% patient_level_ovarian$sample_id)


# data frame with clinical characteristics where each row is a different cell
ovarian_df <- full_join(patient_level_ovarian, cell_level_ovarian, by = "sample_id") %>%
  mutate(sample_id = as.numeric(as.factor(sample_id))) %>%
  filter(tissue_category != "Glass") %>%
  dplyr::select(sample_id, cell_id, tissue_category, x = cell_x_position, y = cell_y_position,
                everything()) %>%
  select(-tma, -diagnosis, -grade)

rm(oc, assays_slot, intensities_df, nucleus_intensities_df, membrane_intensities_df, colData_df, spatialCoords_df, patient_level_ovarian, cell_level_ovarian)
```

The resulting dataframe, `ovarian_df`, contains information on marker intensities, cell X and Y position, cell phenotypes, and patient-level characteristics.

### Exploratory analysis


```{r, eval = FALSE}
dim(ovarian_df)

head(ovarian_df)

glimpse(ovarian_df)
```


Below we plot the cells for a single subject. In this dataset there is one image for each subject, and the image comes from a tumor microarray (TMA).

```{r}
id = sample(ovarian_df$sample_id, 1)

ovarian_df %>%
  filter(sample_id == id) %>%
  ggplot(aes(x, y)) +
  geom_point(aes(color = tissue_category), size = 0.1)


```

Now we plot the images (below) for multiple subjects. In this dataset, multiple TMA cores (each TMA core is frome a different subject) were placed on the same slide, and then that slide was imaged. X and Y locations in this data correspond to position on the slide for a given patient sample. 

```{r}
ovarian_df %>%
  filter(sample_id < 10) %>%
  ggplot(aes(x, y)) +
  geom_point(aes(color = tissue_category), size = 0.1)
```


## Lung data

Load the non small cell lung carcinoma data and inspect the `SpatialExperiment` object. This dataset has 3-4 images per patient, representing different ROIs from a tissue slice. In this dataset, each patient was imaged on a separate slide.

```{r}
# load lung data
lung <- HumanLungCancerV3()

# check object
lung
```

### Converting to dataframe 

Code below converts the lung cancer dataset from a `SpatialExperiment` object to a `data.frame`.

```{r}
## Assays slots
assays_slot <- assays(lung)
intensities_df <- assays_slot$intensities
nucleus_intensities_df<- assays_slot$nucleus_intensities
rownames(nucleus_intensities_df) <- paste0("nucleus_", rownames(nucleus_intensities_df))
membrane_intensities_df<- assays_slot$membrane_intensities
rownames(membrane_intensities_df) <- paste0("membrane_", rownames(membrane_intensities_df))

# colData and spatialData
colData_df <- colData(lung)
spatialCoords_df <- spatialCoords(lung)

# clinical data
patient_level_lung <- metadata(lung)$clinical_data 

cell_level_lung <- as_tibble(cbind(colData_df, 
                                     spatialCoords_df,
                                     t(intensities_df),
                                     t(nucleus_intensities_df),
                                     t(membrane_intensities_df))
                               )   %>%
  dplyr::rename(cd19 = cd19_opal_650,
                cd3 = cd3_opal_520, 
                cd14 = cd14_opal_540,
                cd8 = cd8_opal_620,
                hladr = hladr_opal_690,
                ck = ck_opal_570) %>%
  dplyr::select(cell_id:slide_id, sample_id:dapi,
                entire_cell_axis_ratio:entire_cell_area_square_microns, contains("phenotype"))



# data frame with clinical characteristics where each row is a different cell
lung_df <- full_join(patient_level_lung, cell_level_lung, by = "slide_id") %>%
  #mutate(slide_id = as.numeric(as.factor(slide_id))) %>%
  dplyr::select(image_id = sample_id, patient_id = slide_id, 
                cell_id, x = cell_x_position, y = cell_y_position,
                everything())

rm(lung, assays_slot, intensities_df, nucleus_intensities_df, membrane_intensities_df, colData_df, spatialCoords_df, patient_level_lung,
   cell_level_lung)
```

The resulting dataframe, `lung_df`, contains information on marker intensities, cell X and Y position, cell phenotypes, and patient-level characteristics.

### Exploratory analysis

Below shows images from a single subject. Each image represents a different ROI from the same tissue slice from a single subject. There are 3-6 ROIs for each subject. 

```{r}
id = sample(lung_df$patient_id, 1)

lung_df %>%
  filter(patient_id == id) %>%
  ggplot(aes(x, y)) +
  geom_point(aes(color = tissue_category), size = 0.1) +
  facet_wrap(~image_id)

```


## Data saving

Processed data are saved as `.rda` objects and can also be downloaded directly from the short course website. For the rest of the workshop we will be working with this processed data.


```{r, eval = FALSE}
# save data sets
save(ovarian_df, file = here::here("Data", "ovarian.RDA"))
save(lung_df, file = here::here("Data", "lung.RDA"))

# load processed ovarian data
load(url("https://github.com/julia-wrobel/MI_tutorial/raw/main/Data/ovarian.RDA"))

# load processed lung data
load(url("https://github.com/julia-wrobel/MI_tutorial/raw/main/Data/lung.RDA"))
```



## References

Original citations for the two datasets in the `VectraPolarisData` package, and the package vignette which includes a data dicitonary:


* [Vignette for VectraPolarisData](https://bioconductor.org/packages/release/data/experiment/vignettes/VectraPolarisData/inst/doc/VectraPolarisData.html)
* [Paper for ovarian cancer data](https://aacrjournals.org/mcr/article/19/12/1973/675069/The-Spatial-Context-of-Tumor-Infiltrating-Immune)
* [Paper for non-small cell lung carcinoma data](https://pubmed.ncbi.nlm.nih.gov/34048945/)


More resources for `SpatialExperiment`:

* [Orchestrating Spatially-Resolved Transcriptomics Analysis with Bioconductor](https://lmweber.org/OSTA-book/index.html)
* [SpatialExperiment: infrastructure for spatially-resolved transcriptomics data in R using Bioconductor](https://pubmed.ncbi.nlm.nih.gov/35482478/)