---
title: "Tools and software for functional data analysis of multiplexed imaging data"
output:
  html_document: 
    toc: true
    toc_float: true
---   

```{r, echo = FALSE, message = FALSE}
library(mxfda)
library(tidyverse)
library(patchwork)

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 8,
  fig.height = 4,
  fig.path = "figs", 
  message = FALSE,
  out.width = "90%"
)

theme_set(theme_bw() + theme(legend.position = "bottom"))
```



## Slide Deck

<iframe class="speakerdeck-iframe" frameborder="0" src="https://speakerdeck.com/player/3710e7390c4842afb3f85c22e729ecde" title="MI_fda" allowfullscreen="true" style="border: 0px; background: padding-box padding-box rgba(0, 0, 0, 0.1); margin: 0px; padding: 0px; border-radius: 6px; box-shadow: rgba(0, 0, 0, 0.2) 0px 5px 40px; width: 100%; height: auto; aspect-ratio: 560 / 314;" data-ratio="1.78343949044586"></iframe>

## Overview

This code briefly showcases the use of the `mxfda` package, an `R` package for performing functional data analysis on spatial summary functions from single-cell spatial proteomics data. 


Install `mxfda` package from GitHub:

```{r, eval = FALSE}
#install devtools if not available
if (!require("devtools", quietly = TRUE))
  install.packages("devtools")

#install from github
devtools::install_github("julia-wrobel/mxfda")
```

First we load libraries used in this tutorial.

```{r, eval = FALSE}
library(mxfda)
library(tidyverse)
library(patchwork)
```

Next we load multiplex imaging data. This is a segmented and phenotyped multiplex immunoflurescence dataset of tumor microarray images of high grade serous ovarian cancer, used in [this study](https://aacrjournals.org/mcr/article/19/12/1973/675069/The-Spatial-Context-of-Tumor-Infiltrating-Immune). The full data are part of the `VectraPolarisData` package. 

```{r}
# load processed ovarian cancer data from GitHub
load(url("https://github.com/julia-wrobel/MI_tutorial/raw/main/Data/ovarian.RDA"))
```

```{r}
ovarian_df <- ovarian_df %>%
  filter(tissue_category == "Tumor") 

dim(ovarian_df)
```


## Generate MxFDA object

The `mxfda` package uses an S4 object of class `mxFDA`. These objects are created with the `make_mxfda()` function.

Slots in the `mxFDA` object are designated as follows:

- `Metadata` - stores sample specific traits that may be used as covariates when fitting models
- `Spatial` - a data frame of cell level information (x and y spatial coordinates, phenotype, etc.) that can be used to calculate spatial summary functions
- `subject_key` - a character string for the column in the metadata that denotes the unique subject id
- `sample_key` - a character string for the column in the metadata that denotes the unique sample id. Note that there may be multiple samples per subject, and this id links the metadata and spatial data for each multiplex image sample
- `univariate_summaries` and `bivariate_summaries` - lists of spatial summary functions either imported with `add_summary_function()` or calculated with `extract_summary_functions()`
- `functional_pca` - list of results from functional principle component analysis 
- `functional_mpca` - list of results from multilevel functional principle component analysis
- `functional_cox` - list of functional cox models that have been fit
- `functional_mcox` - list of functional mixed cox models that have been fit

Below, we convert the ovarian cancer data to an MxFDA object.

```{r}
clinical = ovarian_df %>%
  select(sample_id, age = age_at_diagnosis, survival_time, death, stage_bin) %>%
  distinct()

spatial = ovarian_df %>%
  select(-age_at_diagnosis, -survival_time, -death, -stage_bin)

mxf = make_mxfda(metadata = clinical,
                         spatial = spatial,
                         subject_key = "sample_id",
                         sample_key = "sample_id")

rm(ovarian_df, clinical, spatial)
```

Check out the new MxFDA object:

```{r}
summary(mxf)
```

```{r}
class(mxf)
```

## Estimate spatial summary functions

You can use the `mxfda` package to calculate univariate and bivariate Ripley's K and nearest neighbor G-functions, or you can input your own spatial summary functions. Below we calculate univariate Ripley's K for immune cells in the dataset.

```{r, warning = FALSE}
mxf = extract_summary_functions(mxf,
                                extract_func = extract_univariate,
                                summary_func = Kest,
                                r_vec = seq(0, 200, by = 1),
                                edge_correction = "iso",
                                markvar = "immune",
                                mark1 = "immune")
```

Running this code will calculate univariate Ripley's K function for immune cells in each image, and will store these spatial summary functions in the `univariate_summaries` slot of the S3 `mxf` object. To access this slot and view the extracted summary functions, type: 

```{r}
mxf@univariate_summaries$Kest
```

The summaries are returned as a dataframe. The variable `sumfun` is the estimated summary function value, `csr` is the theoretical value under complete spatial randomness, and `fundiff` = `sumfun`-`csr`; in downstream analysis we will use the `fundiff` covariate. 

Next, plot the spatial summary functions we just estimated.

```{r}
plot(mxf, y = "fundiff", what = "uni k") +
  geom_hline(yintercept = 0, color = "red", linetype = 2)
```

Values above the dotted red line are interpreted as having more spatial clustering among immune cells at radius $r$ than would be expected under complete spatial randomness. 

## FPCA

Below we run FPCA on these K-function curves, using the `run_fpca()` function, and store these results in the `mxf` object.

```{r}
mxf <- run_fpca(mxf, 
                metric = "uni k", 
                r = "r", 
                value = "fundiff",
                pve = .99)
```

Note that the summary of this object now shows the number of functional principal components (FPCs) that have been calculated:

```{r}
mxf
```


Below we plot the first two functional principal components from our FPCA analysis of the K functions.

```{r fpc_plots, fig.width = 10}
p1 = plot(mxf, what = 'uni k fpca', pc_choice = 1)
p2 = plot(mxf, what = 'uni k fpca', pc_choice = 2)

(p1 + p2)
```

The black line in each plot is the mean of the data. In these plots, the blue and red lines represent one standard deviation of the FPC added (blue) or subtracted (red) from the population mean. 

## Functional Cox regression

Linear functional Cox regression

```{r}
mxf <- run_fcm(mxf, model_name = "fit_lfcm",
               formula = survival_time ~ age, 
               event = "death",
               metric = "uni k", 
               r = "r", 
               value = "fundiff",
               afcm = FALSE)
```

```{r}
summary(extract_model(mxf, 'uni k', 'fit_lfcm'))
```

